Empirical studies citing Lakens' (2017) equivalence primer

### 2017-11-08 manuscript.Rmd
Major updates to manuscript
* Copied text from Google doc (08-11-17)
* Removed copy-pasted code in examples and figure generation
* Included source commands for generating example code and figures, to avoid errors when updating code outside of manuscript.

Psychologists should be able to falsify predictions. A common prediction in psychological research is that an effect exists in the population that differs from zero. For example, we might predict that priming American Asian women with their Asian identity will increase their performance on a math test compared to women who are primed with their female identity. To be able to falsify this hypothesis, and design a study that allows for strong inferences [@platt_strong_1964], it is important to specify which test result would *disprove* the hypothesis.

An equivalence test can be used to test whether the observed effect is surprisingly small, assuming a meaningful effect exists in the population. The test is a simple variation of the widely used null hypothesis significance tests (NHST). To understand the idea behind equivalence tests, it is useful to realize the null hypothesis we test against can be any numerical value. When we compare two groups, we often test whether the difference between these groups is zero, but we may sometimes want to test against other values than zero. Imagine a researcher who is interested in voluntary participation in a national program to train young infants' motor skills. The researcher wants to test whether more boys than girls are brought into the program by their parents. One could test whether the difference between boys and girls is zero. However, this would ignore the fact that the human sex ratio of newborn boys to girls is not exactly 1:1, and we should not expect 50% of participants to be boys. On average, 103 boys are born for every 100 girls (CIA, 2017), so approximately 50.74% of applicants should be boys, and 49.26% should be girls (with 0.5074/0.4926≈1.03). If boys and girls were exactly equally likely to be brought into the program by their parents, we would not expect a difference of zero, but 1.5% more boys. This can be used as a null hypothesis: Rather than testing against a difference in proportions of 0, the researcher tests against 0.015.

Even though 0.015 might be a more sensible value for a null hypothesis, our researcher knows that the true sex ratio in their population will differ slightly from that in the entire world. Rather than testing a point null hypothesis (setting $H_0$ to one specific value), the researcher decides to define a range of values around the difference in proportions of 0.015 that can be considered trivially small, even when the true ratio in the population is not exactly 0, or even 0.015.  The researcher can for example test if the difference is smaller than -0.005, or larger than 0.035. This test against two bounds, with $H_0$ as a range rather than one value, is known as a *minimal effects test* (Murphy, Myors, & Wolach, 2014).(Murphy et al. 2014)

However, even if the researcher would consider any difference in percentages smaller than $-0.5 \%$ or larger than $3.5 \%$ a rejection of the null hypothesis, not every difference is large enough to matter in practice. As long as gender differences in the participation rate are not too extreme, there is no need to spend money to address the gender imbalance in the voluntary participation rate, and all money can be spent on the core training program. To test whether the gender difference in participants is *not large enough to matter, the researcher can perform an *equivalence test*. Equivalence tests examine whether the presence of gender differences that are large enough to matter can be rejected. After an extensive discussion with experts, the researcher decides that as long as the difference in proportions is not larger than 6% the researcher can act as if the gender difference is too small to intervene. Given an expected true difference in the population of 0.015, the researcher will tests if the observed difference in the data is smaller than 0.075, and larger than -0.055. If differences more extreme than both these boundary values can be rejected in two one-sided tests, the researcher will conclude statistical equivalence, and no money will be spent on addressing a gender difference in participation.

One can conclude statistical equivalence whenever a 90% confidence interval[a][b][c][d][e][f] around the observed estimate does not contain the smallest effect size of interest (SESOI[g]). Figure 1 illustrates the possible conclusions we can draw from null hypothesis, minimal effect, and equivalence tests.

Even though equivalence tests are just a small variation of traditional frequentist null-hypothesis significance tests (and test against the SESOI, instead of a value of 0), their use was limited until the availability of user-friendly software to perform the calculations [@lakens_equivalence_2017]. In this article, we provide several examples of equivalence tests which illustrate the procedure. Furthermore, we discuss different approaches to determining the SESOI for psychological research, and provide detailed reproducible examples of how to perform power analyses when designing equivalence tests, and statistical re-analyses of published psychology experiments.

@@

Equivalence tests are performed against a value that is considered the smallest effect size of interest (SESOI). The SESOI can sometimes be based on just noticeable differences, which can be objectively determined. Most often, however, it is a subjective decision that varies across individuals and time. The SESOI is ideally based on a cost-benefit analysis. Since both costs and benefits are necessarily subjective, the SESOI will depend on the researcher who designs the study. The goal of setting a SESOI is to clearly justify why designing a study that has a high probability of rejecting effects larger than a specified value contributes to our knowledge base. A SESOI should be chosen such that inferences based on it answer a meaningful question.

@@

An objectively determined SESOI should be based on quantifiable theoretical predictions, such as computational models. Sometimes, the only theoretical prediction is that an effect should be noticeable. In such circumstances, the SESOI can be set based on just noticeable differences. For example, Burriss and colleagues [-@burriss_changes_2015] examined whether women displayed an increase in redness in the face during the fertile phase of their ovulatory cycle. The hypothesis was that a slightly redder skin signals greater attractiveness and physical health, and sending this signal to men yields an evolutionary advantage. This hypothesis requires that the increase in redness can be detected with the naked eye by men. They collected data from 22 women and showed that there was indeed an increase in redness of the facial skin of woman during their fertile period. However, this increase was not large enough to be noticeable with the naked eye by men, thus falsifying their hypothesis. Because the just noticeable difference in redness of the skin can be measured, it is possible to objectively establish the SESOI.

@@

@@

We distinguish between three categories of subjective justifications for SESOI. First, researchers can use benchmarks. For example, one might set the SESOI to a standardized effect size of d = 0.5[h][i][j][k], which would allow one to reject effect as large or larger than a 'medium' effect size (Cohen, 1988). Similarly, effect sizes smaller than a Cohen's d of 0.1 are sometimes considered trivially small (Maxwell, Lau, & Howard, 2015). Relying on a benchmark is the weakest possible justification of a SESOI, and should be avoided.

@@

Second, researchers can determine the SESOI based on previous work in the literature. Ideally, researchers would specify their SESOI in studies they report, but this is not yet common practice. It is thus up to researchers who build on earlier work to decide which effect size is too small to be meaningful, given an earlier study. Simonsohn (2015) recently proposed to set the SESOI to the effect size that an earlier study on the same effect would have 33% power to detect based on the sample size. This is called the 'small telescopes' approach. For example, consider a study where `r N` participants answered a question, which was analyzed with an one-sample t-test. For a two-sided test with an alpha of `r alpha` this test had 33% power to detect a d = `r round(pwr.t.test(n=N,sig.level=alpha,power=0.33,type="one.sample",alternative="two.sided")$d,3)`.

Another justifiable choice for the SESOI we would like to propose here is to use the smallest observed effect size that could have been statistically significant in the original study. In other words, we decide that effects that could not have yielded a *p* < 0.05 in the original study (if this is your preferred alpha threshold) will not be considered meaningful in the replication either, even if effects of this size are statistically significant in the replication study. Based only on the alpha level and the sample size, we can calculate the critical test value (e.g., *t*, *F*, *Z*). This critical test value can also be transformed to a standardized effect size (e.g., $d = t \sqrt { \frac { 1} { n _ { 1} } + \frac { 1} { n _ { 2} } }$), which can thus be interpreted as a *critical effect size*. All observed effect sizes smaller than the critical effect size would not be statistically significant in an original study, given the alpha and sample size of that study. By setting the SESOI to the critical effect size, an equivalence test can reject all observed effect sizes that could have been detected in an earlier study.

@@

Third[l][m], researchers can set the SESOI based on the resources they have available. The amount of data you can collect limits the inferences you can make. Given a an alpha level and a sample size, researchers can calculate the smallest effect size that one has sufficient power to detect.[n] This approach is suggested here to provide researchers with a way to think about setting the SESOI for a study in a line of research where no quantifiable predictions are made, and where no previous studies have been performed. Observing statistical equivalence for a SESOI based on this approach allows one to reject effects that can be reliably detected with the given sample size. For example, a researcher who plans to perform a two-sided one-sample *t*-test using an alpha of 5%, based on data from 100 observations, has 90% power to detect an effect of d = 0.364. Using equivalence bounds of d = -0.364 and d = 0.364 in an equivalence test allows one to reject effects the study had high statistical power to detect. In this case, concluding equivalence would suggest one can reject effects that can reasonably be detected with N = 100. Whether or not this is interesting depends in part on how typical the chosen sample size is for a research area. One should not expect an equivalence test based on 15 observations testing against a SESOI of d = -1 and d = 1 to substantially increase our knowledge, given that most effects in psychology are substantially smaller than these bounds. But by transparently reporting the effects one can detect and reject, based on the study design, researchers can communicate the information their study contributes. When there are no previous studies to build on, and no theoretical predictions, specifying a SESOI based on the alpha level and a sample size can provide a starting point for a discussion on what a reasonable SESOI is. Nevertheless, the value of setting a SESOI based on this procedure is limited, and the equivalence test does not reject values that are theoretically interesting.

```{r}
p <- 0.05
N <- 100
crit_d <- abs(qt(p/2, (N*2)-2))/sqrt(N/2)
```

@@

```{r include = FALSE}
library(TOSTER)

# original study: Banerjee et al., study 1
orig.m1 <- 4.71
orig.m2 <- 5.3
orig.sd1 <- 0.85
orig.sd2 <- 0.97
orig.d <- 0.65
orig.t <- 2.03
orig.p <- 0.049
orig.N <- 40 #group size unknown, therefore equal n assumed
orig.df <- orig.N-2

# replication study: Brandt et al., study 1
rep.m1 <- 4.7857
rep.m2 <- 4.6569
rep.sd1 <- 1.0897
rep.sd2 <- 1.1895
rep.n1 <- 49
rep.n2 <- 51
rep.t <- 0.56
rep.p <- .574
rep.df <- rep.n1 + rep.n2 - 2
rep.d <- 0.11

# Calculate critical effect (d): smallest effect orig. study had the power to detect
t.crit = qt(1-.05/2, (orig.N-2))
d.crit = t.crit*sqrt((1/(orig.N/2)) + 1/(orig.N/2))

Brandt <- TOSTtwo(m1=rep.m1, m2=rep.m2, sd1=rep.sd1, sd2 = rep.sd2, n1 = rep.n1, n2=rep.n2, low_eqbound_d = -d.crit, high_eqbound_d = d.crit, var.equal = FALSE)
```

Banerjee, Chatterjee, & Sinha (2012) reported that participants who had been asked to describe an unethical deed from their past judged the room to be darker than participants who had been asked to describe an ethical deed ($M_{unethical}= `r orig.m1`$, ${SD}_{unethical}= `r orig.sd1`$, $M_{ethical}=`r orig.m2`$, ${SD}_{ethical}=`r orig.sd2`$, $t(`r orig.df`)= `r orig.t`$, $p= `r orig.p`$, $d= `r orig.d`$). A close replication by Brandt, IJzerman, & Blanken (2014) found no significant effect ($t(`r rep.df`)=`r rep.t`$, $p=`r rep.p`$, $d=`r rep.d`$).  The smallest effect size that could have been statistically significant in the original study is $d_{crit}= `r round(t.crit, 2)` \sqrt { \frac { 1} {`r orig.N/2`} + \frac { 1} {`r orig.N/2`} }=`r round(d.crit, 2)`$. Using this as our SESOI for a TOST with Welch's *t*-test for independent samples --- resulting in equivalence bounds of ${\Delta}_L=`r round(-d.crit, 2)`$ and ${\Delta}_U=`r round(d.crit, 2)`$ --- we indeed find that the effect reported by the replication study is statistically equivalent, $t(`r round(Brandt$TOST_df, 2)`)=`r round(min(Brandt$TOST_t1, Brandt$TOST_t2), 2)`$, $p=`r round(max(Brandt$TOST_p1, Brandt$TOST_p2), 3)`$.

```{r, fig.width=6}
TOSTtwo(m1=rep.m1, m2=rep.m2, sd1=rep.sd1, sd2 = rep.sd2, n1 = rep.n1, n2=rep.n2, low_eqbound_d = -d.crit, high_eqbound_d = d.crit, var.equal = FALSE)
```

[^1]: This will typically, although not always, correspond to the effect size the study had $50\%$ power to detect[o]. This procedure will thus result in equivalence bounds that are substantially larger than the ones obtained using the small telescopes approach, which gives the effect size a study had $33\%$ power to detect[p][q][r][s][t][u][v][w].

@@

```{r, fig.width=6}
library("TOSTER")
TOSTmeta(ES = 0.06, se = 0.003, low_eqbound_d=-0.1, high_eqbound_d=0.1, alpha=0.005)
TOSTmeta(ES = 0.04, se = 0.002, low_eqbound_d=-0.1, high_eqbound_d=0.1, alpha=0.005)
TOSTmeta(ES = -0.01, se = 0.002, low_eqbound_d=-0.1, high_eqbound_d=0.1, alpha=0.005)
TOSTmeta(ES = -0.01, se = 0.002, low_eqbound_d=-0.1, high_eqbound_d=0.1, alpha=0.005)
TOSTmeta(ES = -0.01, se = 0.002, low_eqbound_d=-0.1, high_eqbound_d=0.1, alpha=0.005)
TOSTmeta(ES = -0.02, se = 0.002, low_eqbound_d=-0.1, high_eqbound_d=0.1, alpha=0.005)
TOSTmeta(ES = -0.02, se = 0.002, low_eqbound_d=-0.1, high_eqbound_d=0.1, alpha=0.005)
TOSTmeta(ES = -0.01, se = 0.003, low_eqbound_d=-0.1, high_eqbound_d=0.1, alpha=0.005)
TOSTmeta(ES = 0.04, se = 0.003, low_eqbound_d=-0.1, high_eqbound_d=0.1, alpha=0.005)
TOSTmeta(ES = 0.06, se = 0.003, low_eqbound_d=-0.1, high_eqbound_d=0.1, alpha=0.005)
```
Hyde, Lindberg, Linn, Ellis, and Williams (2008) report that effect sizes for gender differences in mathematics tests across the 7 million students in the US represent trivial differences, where the authors specify a trivial difference as an effect size smaller than d = 0.1. For example, in grade 3 the difference is d = 0.04, with a standard error of 0.002. When we perform equivalence tests on the meta-analytic effect sizes of IQ difference for grades 2 to 11 (using an alpha level of 0.005 to correct for multiple comparisons), using equivalence bounds of d = -0.1 and d = 0.1, we see that indeed, all effect size estimates are measured with such high precision, we can conclude that they fall within the equivalence bound of d = -0.1 and d = 0.1, and can be considered trivially small, according to the bounds set by the authors. However, note that all of the effects are also statistically different from zero, as one might expect when there is no random assignment to conditions, and samples sizes are huge. This is one way in which the use of equivalence tests improve hypothesis testing procedures in science, by distinguishing between statistical significance and practical significance[x].

## Example 3: Not Statistically Equivalent and Not Statistically Different

@@

Using the TOSTtwo function in the TOSTER package in R, we can then calculate the TOST by inserting the summary statistics from Moon & Roeder's (2014) data, assuming an alpha of 0.05, and using +/- the SESOI of 0.0625 as our equivalence bounds.[y][z][aa][ab][ac]

```{r echo=FALSE}

#  We first load the relevant datafile into R
moon.data        <- read.csv("NSNEQ_Moon_Example_data.csv")

#  We then calculate the summary statistics from the data that we need in the TOST calculation below
mean.asian   <- mean(moon.data$accuracy[moon.data$identity_salience == 1])
mean.control <- mean(moon.data$accuracy[moon.data$identity_salience == 2])
sd.asian         <- sd(moon.data$accuracy[moon.data$identity_salience == 1])
sd.control   <- sd(moon.data$accuracy[moon.data$identity_salience == 2])
n.asian          <- sum(moon.data$identity_salience == 1)
n.control        <- sum(moon.data$identity_salience == 2)

#  We can now run the TOST!
#  Performing the TOST using the 'TOSTtwo.raw' function allows us to specify equivalence bounds in raw values
tost <- TOSTER::TOSTtwo.raw(m1  = mean.asian,
                                m2  = mean.control,
                                sd1 = sd.asian,
                                sd2 = sd.control,
                                n1  = n.asian,
                                n2  = n.control,
                                low_eqbound  = -0.0625,
                                high_eqbound =  0.0625,
                                alpha = .05)
```

We find that the TOST is non-significant, t(`r round(tost$TOST_df, 2)`) = `r round(min(abs(c(tost$TOST_t1, tost$TOST_t2))), 2)`, p = `r round(max(tost$TOST_p1, tost$TOST_p2), 2)`. Thus, we cannot reject the possibility that the true effect really is larger than our SESOI.[^footnote]

[^footnote]:[ad] In this example, the bounds of the TOST was set based on a raw effect of 6.25%. We could have converted this raw effect into a standardized effect (Cohen's d) for the current sample, and we would have gotten the exact same result, only then in Cohen's d (equivalence bounds = +/- `r round(0.0625 / sqrt((sd.asian^2 + sd.control^2) / 2), 2)`). Since this is a replication of a previous study, we could also have chosen calculate the Cohen's d that corresponds to the raw SESOI in the *original* study, and used this as the bounds for the replication. Now we would no longer be answering a question about a raw effect independent of the samples. Instead, we are asking whether we in the replication can reject standardized effects larger than a certain standardized effect in the original study. Since the standard deviations of the original study were only slightly higher than in the replication in this case, Cohen's d, and thus the bounds, would also be only sligthly smaller (bounds = +/- `r round(0.0625 / sqrt((0.17^2 + 0.20^2) / 2), 2)`). As standard deviations of the original study compared to the replication becomes larger, the bounds become more narrow.

## Example 4: Statistically Inferior [ae][af]and Not Statistically Different

>>>>

Psychologists should be able to falsify predictions. A common prediction in psychological research is that an effect exists in the population that differs from zero. For example, we might predict that priming American Asian women with their Asian identity will increase their performance on a math test compared to women who are primed with their female identity. To be able to falsify this hypothesis, and design a study that allows for strong inferences (Platt, 1964), it is important to specify which test result would *disprove* the hypothesis.

An equivalence test can be used to test whether the observed effect is surprisingly small, assuming a meaningful effect exists in the population. The test is a simple variation of the widely used null hypothesis significance tests (NHST). To understand the idea behind equivalence tests, it is useful to realize the null hypothesis we test against can be any numerical value. When we compare two groups, we often test whether the difference between these groups is zero, but we may sometimes want to test against other values than zero. Imagine a researcher who is interested in voluntary participation in a national program to train young infants' motor skills. The researcher wants to test whether more boys than girls are brought into the program by their parents. One could test whether the difference between boys and girls is zero. However, this would ignore the fact that the human sex ratio of newborn boys to girls is not exactly 1:1, and we should not expect 50% of participants to be boys. On average, 103 boys are born for every 100 girls (CIA, 2017), so approximately 50.74% of applicants should be boys, and 49.26% should be girls (with $0.5074/0.4926 \approx 1.03$). If boys and girls were exactly equally likely to be brought into the program by their parents, we would not expect a difference of zero, but 1.5% more boys. This can be used as a null hypothesis: Rather than testing against a difference in proportions of 0, the researcher tests against 0.015.

Even though 0.015 might be a more sensible value for a null hypothesis, our researcher knows that the true sex ratio in their population will differ slightly from that in the entire world. Rather than testing a point null hypothesis (setting $H_0$ to one specific value), the researcher decides to define a range of values around the difference in proportions of 0.015 that can be considered trivially small, even when the true ratio in the population is not exactly 0, or even 0.015.  The researcher can for example test if the difference is smaller than -0.005, or larger than 0.035. This test against two bounds, with $H_0$ as a range rather than one value, is known as a *minimal effects test* (Murphy, Myors, & Wolach, 2014).

However, even if the researcher would consider any difference in percentages smaller than $-0.5 \%$ or larger than $3.5 \%$ a rejection of the null hypothesis, not every difference is large enough to matter in practice. As long as gender differences in the participation rate are not too extreme, there is no need to spend money to address the gender imbalance in the voluntary participation rate, and all money can be spent on the core training program. To test whether the gender difference in participants is *not* large enough to matter, the researcher can perform an *equivalence test*. Equivalence tests examine whether values below and above a prespecified range can be rejected. Here, we would test if the presence of gender differences that are large enough to matter can be rejected. After an extensive discussion with experts, the researcher decides that as long as the difference in proportions is not larger than 6% the researcher can act as if the gender difference is too small to intervene. Given an expected true difference in the population of 0.015, the researcher will test if the observed difference in the data is smaller than -0.055, and larger than 0.075. If differences more extreme than both these boundary values can be rejected in two one-sided tests, the researcher will conclude statistical equivalence, and no money will be spent on addressing a gender difference in participation.

For an alpha level of 5%, we can conclude statistical equivalence whenever a 90% confidence interval around the observed estimate does not contain the smallest effect size of interest (SESOI). A 90% confidence interval is used because one-sided tests are performed, and it is not necessary to control for multiple comparisons when performing two one-sided tests because both one-sided tests need to be statistically significant to conclude equivalence. Note that although we will use an alpha of 5% in all examples because this alpha level was used in the papers we reanalyze, researchers should carefully decide upon a desired Type 1 error rate, and justify their choice (Lakens et al., 2017). Figure 1 illustrates the possible conclusions we can draw from null hypothesis, minimal effect, and equivalence tests. Although thinking about equivalence tests in terms of the values that can be rejected based on a confidence interval might be the easiest conceptualization, for any given study, you can also think of an equivalence test as determining whether effect sizes or test statistics are closer to zero than some critical, or even whether the *p*-value for a NHST is larger than some *p*-value bound.

```{r echo=FALSE, fig.width=5, fig.height=7}
#  Plot H0 and H1 regions for different significance tests
source("H0H1_plot.R")
```

Figure caption: Figure 1. Illustration of null hypotheses ($H_0$) and alternative hypotheses ($H_1$) for different types of significance tests. **A)** Equivalence test: Tests the hypothesis ($H_0$) that an effect is smaller than $\Delta_{L}$ *or* larger than $\Delta_{U}$. **B)** NHST: Tests the hypothesis ($H_0$) that an effect is equal to 0. **C)** Minimal effects test: Tests the hypothesis ($H_0$) that an effect is larger than $\Delta_{L}$ *and* smaller than $\Delta_{U}$. **D)** Inferiority test: Tests the hypothesis ($H_0$) that an effect is larger than $\Delta$.

Even though equivalence tests are just a small variation of traditional frequentist null-hypothesis significance tests (and test against the SESOI, rather than against 0), their use in psychology was limited, perhaps due to the lack of user-friendly software to perform the calculations in the past [@lakens_equivalence_2017]. In this article, we provide several examples of equivalence tests which illustrate the procedure in free software (R, a spreadsheet, and in Jamovi). Furthermore, we discuss different approaches to determining the SESOI for psychological research, and provide detailed reproducible examples of how to perform power analyses when designing equivalence tests, and statistical re-analyses of published psychology experiments.

@@

Equivalence tests are performed against a value that is considered the smallest effect size of interest (SESOI). Although in the examples here, we use symmetrical bounds around 0 (e.g., -0.3, 0.3), based on for example an explicit cost-benefit analysis, one might use asymmetric bounds (e.g., -0.2, 0.3). The SESOI can sometimes be determined objectively, for example based on just noticeable differences . Most often, however, it is a subjective decision that varies across researchers, fields, and time. In lieu of objective justifications, the SESOI should ideally be based on a cost-benefit analysis. Since both costs and benefits are necessarily subjective, the SESOI will depend on the researcher who designs the study. The goal of setting a SESOI is to clearly justify why designing a study that has a high probability of rejecting effects larger than a specified value contributes to our knowledge base. A SESOI should be chosen such that inferences based on it answer a meaningful question.

@@

An objectively determined SESOI should be based on quantifiable theoretical predictions, such as computational models. Sometimes, the only theoretical prediction is that an effect should be noticeable. In such circumstances, the SESOI can be set based on just noticeable differences. For example, Burriss and colleagues [-@burriss_changes_2015] examined whether women displayed an increase in redness in the face during the fertile phase of their ovulatory cycle. The hypothesis was that a slightly redder skin signals greater attractiveness and physical health, and sending this signal to men yields an evolutionary advantage. This hypothesis presupposes that men can detect the increase in redness with the naked eye. Burriss and colleagues collected data from 22 women and showed that there was indeed an increase in redness of the facial skin during their fertile period. However, this increase was not large enough for men to detect with the naked eye, thus falsifying their hypothesis. Because the just noticeable difference in redness of the skin can be measured, it is possible to objectively establish the SESOI.

@@

Norman, Sloan, and Wyrwich (2003) used a wide range of methods to determine the minimally important (or detectable) difference in measures for the quality of life in chronic diseases, and observe a surprisingly consistent minimally important difference of half a standard deviation, or d = 0.5. They make the argument that this bound of approximately half a standard deviation could be related to the limits of human discriminability. Therefore, even though a SESOI of d = 0.5 might seem quite large, it could be a plausible equivalence bound when examining effects that should be noticeable by single individuals.

@@

We distinguish between three categories of subjective justifications for SESOI. First, researchers can use benchmarks. For example, one might set the SESOI to a standardized effect size of d = 0.5, which would allow one to reject effect as large or larger than a 'medium' effect size (Cohen, 1988). Similarly, effect sizes smaller than a Cohen's d of 0.1 are sometimes considered trivially small (Maxwell, Lau, & Howard, 2015). Relying on a benchmark is the weakest possible justification of a SESOI, and should be avoided. Based on a review of 112 meta-analyses, Weber and Popova (2012) conclude that setting a SESOI in communication research to a medium effect size (*r* = 0.3, or d = 0.5) corresponds to rejecting the highest 25% of effect sizes reported in communications research, and Hemphill (2003) suggests that in psychology a SESOI of d = 0.5 would imply rejecting effects as large as the upper 33% of effect sizes reported in the psychological literature.

@@

Second, researchers can determine the SESOI based on previous work in the literature. Ideally, researchers would specify their SESOI in studies they report, but this is not yet common practice. It is thus up to researchers who build on earlier work to decide which effect size is too small to be meaningful, given an earlier study. Simonsohn (2015) recently proposed to set the SESOI to the effect size that an earlier study on the same effect would have had 33% power to detect. With this so-called 'small telescopes' approach, the equivalence bounds are thus determined primarily based on the sample size in the original study. For example, consider a study in which `r N` participants answered a question, and the results were analyzed with a one-sample t-test. For a two-sided test with an alpha of `r alpha`, this test had 33% power to detect an effect of d = `r round(pwr.t.test(n=N,sig.level=alpha,power=0.33,type="one.sample",alternative="two.sided")$d,3)`. Another example of how previous research can be used to determine the SESOI can be found in Kordsmeyer and Penke (2017), who defined the SESOI based on the mean of effect sizes reported in the literature. Thus, they examined whether their replication study could reject effects as large or larger than had on average been reported in the literature. Given random variation and bias in the literature, a more conservative approach could be to use a lower bound of a confidence interval around the meta-analytic effect size estimate (cf. Perugini, Gallucci, & Constantini, 2014).

Another justifiable choice for the SESOI is to use the smallest observed effect size that could have been statistically significant in a previous study. In other words, we decide that effects that could not have yielded a $p < \alpha$ in an original study will not be considered meaningful in the replication study either, even if they are found to be statistically significant. Based only on the alpha level and the sample size, we can calculate the critical test value (e.g., *t*, *F*, *Z*). This critical test value can be transformed to a standardized effect size (e.g., $d = t \sqrt { \frac { 1} { n _ { 1} } + \frac { 1} { n _ { 2} } }$), which can thus be interpreted as a *critical effect size*[^footnote]. All observed effect sizes smaller than the critical effect size would not have been statistically significant in the original study, given the alpha and sample size of that study. By setting the SESOI to the critical effect size, an equivalence test can reject all observed effect sizes that could have been detected in an earlier study.

[^footnote]: This will typically, although not always, correspond to the effect size the study had $50\%$ power to detect (see Lenth, 2007). This procedure will thus result in equivalence bounds that are substantially wider than the ones obtained using the small telescopes approach, which gives the effect size a study had $33\%$ power to detect.

@@

Third, researchers can set the SESOI based on the sample size they are planning to collect. The amount of data you can collect limits the inferences you can make. Given the alpha level and the sample size, researchers can calculate the smallest effect size that they have sufficient power to detect. This approach can be used when there are no quantitative predictions made by a theory, and no previous studies have been performed, but researchers *can* justify their sample size. In essence, this approach allows researchers to reject the presence of effects that can be reliably detected given a specific sample size. For example, a researcher who plans to perform a two-sided one-sample *t*-test using an alpha of 5%, based on data from 100 observations, has 90% power to detect an effect of d = 0.364. Using equivalence bounds of d = -0.364 and d = 0.364 in an equivalence test allows one to reject effects the study had high statistical power to detect. An equivalence test based on this approach does not answer any theoretical question (after all, the equivalence bounds are not based on any theoretical predictions) but it answers a *resource* question: Can effects be rejected that can be reliably detected given a specific sample size? In this case, concluding equivalence would suggest one can reject effects that can reasonably be detected with a sample size of 100, and that future studies probably need larger sample sizes to examine a specific question. Whether or not the answer such an equivalence test provides is interesting depends on the sample sizes used in the study, and in part on how typical the chosen sample size is for the research area. One should not expect an equivalence test based on 15 observations testing against a SESOI of d = -1 and d = 1 to substantially increase our knowledge, given that most effects in psychology are substantially smaller than these bounds and require larger sample sizes to be detected. By transparently reporting the effects one can detect and reject, based on the study design, researchers can communicate the information their study contributes, and provide a starting point for a discussion about what a reasonable SESOI is. 
```{r}
p <- 0.05
N <- 100
crit_d <- abs(qt(p/2, (N*2)-2))/sqrt(N/2)
```
#Raw versus standardized equivalence bounds

Equivalence bounds can be set in terms of standardized effect sizes, or raw differences. For example, the SESOI might be a raw mean difference of 0.5 points on a 7-point scale, or a Cohen’s d of 0.5. The main difference is that equivalence bounds set in raw differences are independent of the standard deviation, while equivalence bounds set as standardized effects are dependent on the standard deviation (since they are calculated as [FORMULA (X1-X2)/SD]. The observed standard deviation randomly varies across samples. In practice, this means that when you use standardized differences as bounds (e.g., d = 0.35) the equivalence test depends on the standard deviation in your the sample. The equivalence bound for a raw mean difference of 0.5 equals a standardized equivalence bound of d = 0.5 when the standard deviation is 1, but a standardized equivalence bound of d = 1 when the standard deviation is 0.5.

@@

```{r include = FALSE}
source("NSEQ_Brandt_Example.R")

Banerjee, Chatterjee, & Sinha (2012) reported that `r orig.N` participants who had been asked to describe an unethical deed from their past judged the room to be darker than participants who had been asked to describe an ethical deed ($M_{unethical}= `r orig.m1`$, ${SD}_{unethical}= `r orig.sd1`$, $M_{ethical}=`r orig.m2`$, ${SD}_{ethical}=`r orig.sd2`$, $t(`r orig.df`)= `r orig.t`$, $p= `r orig.p`$, $d= `r orig.d`$). A close replication by Brandt, IJzerman, & Blanken (2014) with `r rep.n1+rep.n2` participants found no significant effect ($t(`r rep.df`)=`r rep.t`$, $p=`r rep.p`$, $d=`r rep.d`$). Following the small telescopes approach, we can calculate the effect size the original study had $33\%$ power to detect --- $d = `r round(d.33, 2)`$ --- and use this as our SESOI. If we run a TOST with Welch's *t*-test for independent samples and equivalence bounds of ${\Delta}_L=`r round(-d.33, 2)`$ and ${\Delta}_U=`r round(d.33, 2)`$, we indeed find that the effect reported by the replication study is statistically equivalent, $t(`r round(Brandt$TOST_df, 2)`)=`r round(min(Brandt$TOST_t1, Brandt$TOST_t2), 2)`$, $p=`r round(max(Brandt$TOST_p1, Brandt$TOST_p2), 3)`$.

@@

```{r, fig.width=6}
source("SEQ_Hyde_Example.R")
```
Hyde, Lindberg, Linn, Ellis, and Williams (2008) report that effect sizes for gender differences in mathematics tests across the 7 million students in the US represent trivial differences, where the authors specify a trivial difference as an effect size smaller than d = 0.1. For example, in grade 3 the difference is d = 0.04, with a standard error of 0.002. When we perform equivalence tests on the meta-analytic effect sizes of IQ difference for grades 2 to 11 (using an alpha level of 0.005 to correct for multiple comparisons), using equivalence bounds of d = -0.1 and d = 0.1, we see that indeed, all effect size estimates are measured with such high precision, we can conclude that they fall within the equivalence bound of d = -0.1 and d = 0.1, and can be considered trivially small, according to the bounds set by the authors (figure 2B). However, note that all of the effects are also statistically different from zero, as one might expect when there is no random assignment to conditions, and samples sizes are huge. This is one way in which the use of equivalence tests improve hypothesis testing procedures in science, by distinguishing between statistical significance and practical significance.

## Example 3: Not Statistically Equivalent and Not Statistically Different
```{r include=FALSE}
source("NSNEQ_Moon_Example.R")
```

@@

We then calculate the TOST for a two-sample Welch’s t-test, assuming an alpha of 0.05 (two-sided), and using +/- the SESOI of 0.0625 as our equivalence bounds ($\Delta$).

We find that the TOST is non-significant, t(`r round(Moon$TOST_df, 2)`) = `r round(min(abs(c(Moon$TOST_t1, Moon$TOST_t2))), 2)`, p = `r round(max(Moon$TOST_p1, Moon$TOST_p2), 2)` (figure 2C). Thus, we cannot reject the possibility that the true effect really is larger than our SESOI.[^footnote] When results are neither statistically different from zero nor statistically equivalent, there is insufficient data to draw conclusions. Further studies are needed, which can be analyzed using a (small-scale) meta-analysis. The additional data will narrow the confidence interval around the observed effect, allowing us to reject the null, reject the SESOI, or both. Analogous to the large sample sizes needed to detect small effects, a lot of data is needed to reject the SESOI when very narrow bounds are used for the equivalence test (e.g., $\Delta_{lower}$ = -0.1 and $\Delta_{upper}$ = 0.1).

[^footnote]: In this example, the bounds of the TOST was set based on a raw effect of 6.25%. We could have converted this raw effect into a standardized effect (Cohen's d) for the current sample, and we would have gotten the exact same result, only then in Cohen's d ($\Delta = +/- `r round(0.0625 / sqrt((sd.asian^2 + sd.control^2) / 2), 2)`)$. Since this is a replication of a previous study, we could also have chosen to calculate the Cohen's d that corresponds to the raw SESOI in the *original* study, and used this as the bounds for the replication. Now we would no longer be answering a question about a raw effect independent of the samples. Instead, we are asking whether we in the replication can reject standardized effects larger than a certain standardized effect in the original study. Since the standard deviations of the original study were only slightly higher than in the replication in this case, Cohen's d (and thus the bounds) would also be only slightly smaller ($\Delta = +/- `r round(0.0625 / sqrt((0.17^2 + 0.20^2) / 2), 2)`)$. As standard deviations of the original study compared to the replication become larger, the bounds become more narrow.

## Example 4: Statistically Inferior and Not Statistically Different
```{r include=FALSE}
source("NSEQ_Lynott_example.R")
```

Lynott et al. (2014) conducted a study to investigate the effect of being exposed to physically warm or cold stimuli on subsequent judgments related to interpersonal warmth and prosocial behavior (replicating Williams and Bargh, 2008). They observed that `r selfish.cold * 100`\% of participants who received a cold pack ($n_{cold}=`r n.cold`$) opted to receive a reward for themselves, while `r selfish.hot * 100`\% of participants who received a warm pack ($n_{hot}=`r n.hot`$) did the same. In a z-test for the difference between the proportions, this effect is not statistically significant ($Diff=`r Lynott$dif * 100`$\%, $Z=`r Lynott$NHST_z`$, $p=`r Lynott$NHST_p`$). Having obtained a non-significant result, we are left with the same two  alternatives as in example 3.

In this case, it is not really clear what an objective criterion for the smallest effect size of interest could be. However, since this is a replication, we could decide that we do not care about an effect if the difference is smaller than the smallest effect that the original study could detect. This essentially means that we use the critical z value (~1.96 in a two-tailed test with an alpha of 0.05) as our bounds ($\Delta = +/- round(qnorm(1-0.05/2),2)$). To figure out what difference corresponds to a critical z in the original study, we multiply critical z with the standard error:
